A DPG method for compressible flow problems

### 2012-06-20 docs/Jesse/Proposal/DPGrobustness.tex

@@

\seclab{sec:modelSec}
The remainder of this chapter focuses on the convection-diffusion problem using the abstract theory that we
have discussed so far. In particular, we shall use the DPG method based on the ultra-weak formulation with optimal test functions to solve this model problem and analyze its behavior with respect to $\epsilon$. 
Our goal is to show the robustness of the method with respect to $\epsilon$, and demonstrate its usefulness as  a numerical approach to solving singular-perturbed problems.

@@

\section{Numerical experiments}
\label{sec:num_exp}

@@

\subsection{Eriksson-Johnson model problem}

@@

\subsubsection{Solution with $C_1 = 1, C_{n\neq 1} = 0$}

@@

\subsubsection{Neglecting $\sigma_n$}

@@

\subsubsection{Discontinuous inflow data}

>>>>

The majority of this chapter will focus on the convection-diffusion problem using the abstract theory that we have discussed in the previous chapter. In particular, we shall use the DPG method based on the ultra-weak formulation with optimal test functions to solve this model problem and analyze its behavior as $\epsilon\rightarrow 0$. Our goal is to show the robustness of the method with respect to $\epsilon$, and demonstrate its usefulness as a numerical method for solving singular-perturbed problems.

@@

\seclab{sec:modelSec}

@@

\section{Numerical experiments: Eriksson-Johnson problem}
\label{sec:num_exp}

@@

@@

\subsection{Solution with $C_1 = 1, C_{n\neq 1} = 0$}

@@

\subsection{Neglecting $\sigma_n$}

@@

\subsection{Discontinuous inflow data}

### 2012-06-20 docs/Jesse/Proposal/DPGtheory.tex

Using a variant of the quasi-optimal test norm, numerical results show that the DPG method appears to provide a ``pollution-free" method without phase error for the Helmholtz equation \cite{DPG4}, and analysis of the pollution-free nature of DPG is currently under investigation. Similar results have also been obtained in the context of elasticity \cite{DPGElas} and the linear Stokes equations \cite{Camellia}. On the theoretical side, the quasi-optimal test norm has been shown to yield a well-posed DPG methodology for the Poisson and convection-diffusion equations \cite{analysisDPG}. More recently, this theory has been generalized to show the well-posedness of DPG for the large class of PDEs of Friedrichs' type \cite{Bui-ThanhDemkowiczGhattas11b}.

>>>>

Using a variant of the quasi-optimal test norm, numerical results show that the DPG method appears to provide a ``pollution-free" method without phase error for the Helmholtz equation \cite{DPG4}, and analysis of the pollution-free nature of DPG is currently under investigation. Similar results have also been obtained in the context of elasticity \cite{DPGElas} and the linear Stokes equations \cite{Camellia}. On the theoretical side, the quasi-optimal test norm has been shown to yield a well-posed DPG methodology for the Poisson and convection-diffusion equations \cite{analysisDPG}. More recently, this theory has been generalized to show the well-posedness of DPG under the quasi-optimal test norm for the large class of Friedrichs' type PDEs \cite{Bui-ThanhDemkowiczGhattas11b}.

### 2012-06-20 docs/Jesse/Proposal/NSNotes.tex

For $p=2$, after six refinements, the mesh contained 1410 elements and 187605 global dofs, with a total DPG energy error of $5.684e-7$. Comparatively, for $p=3$, the mesh contained only 444 elements and 97403 global dofs, with a total DPG eneryg error of 4.57681e-7 after five refinements.

>>>>

For $p=2$, after six refinements, the mesh contained 1410 elements and 187605 global dofs, with a total DPG energy error of $.000753$. Comparatively, for $p=3$, the mesh contained only 444 elements and 97403 global dofs, with a total DPG eneryg error of $0.000676$ after five refinements.

### 2012-06-20 docs/Jesse/Proposal/abstract.tex

\author{Jesse Chan}
\address{910 E. 32nd St, Apt.\ 101\\ Austin, Texas 78705}  % Required

\title{Proposal outline/notes for DPG and Compressible Navier-Stokes}

\supervisor
	[Leszek Demkowicz]
	{Robert Moser}

\committeemembers
	[Thomas J. R. Hughes]
	[Committee member 2]
	[Committee member 3]
	{Committee member 4}

\previousdegrees{B.A.}

\oneandonehalfspacequote

\topmargin 0.125in

\makeindex

%%%%%%%%%%%%%%% Start of thesis %%%%%%%%%%%%%%%%%%%

\begin{document}
\copyrightpage          % Produces the copyright page.
\commcertpage           % Produces the Committee Certification
\titlepage

\begin{dedication}
\index{Dedication@\emph{Dedication}}%
Dedicated to \ldots
\end{dedication}

\begin{acknowledgments}		% Optional
\index{Acknowledgments@\emph{Acknowledgments}}%
I wish to thank the multitudes of people who helped me. Time would
fail me to tell of \ldots
\end{acknowledgments}

\utabstract
\index{Abstract}%
\indent
This document has the form of a ``fake'' doctoral dissertation
in order to provide an example of such, but it is actually a
copy of Miguel Lerma's documentation for the Mathematics
Department Computer Seminar of 25 March 1998 updated in July 2001
and following by Craig McCluskey to meet the March 2001
requirements of the Graduate School.

This document and its source file show to write a Doctoral Dissertation using 
\LaTeX{} and the utdiss2 package.

\tableofcontents   % Table of Contents will be automatically
                   % generated and placed here.

%\listoftables      % List of Tables and List of Figures will be placed
%\listoffigures     % here, if applicable.

\chapter{Introduction}

\input{proposalIntro}

\chapter{Range of problems}

\input{problemRange}

\chapter{Discontinuous Petrov-Galerkin: a minimum residual method for linear problems}

\input{DPGtheory}

\chapter{A robust DPG method for convection-diffusion}

\input{DPGrobustness}

\chapter{Proposed work}

\input{NonlinDPG}
\input{BurgersNotes}
\input{NSNotes}
\input{areaReq}

>>>>

\author{Jesse Chan\\Committee: \vspace{-0.3cm}\\Dr. Leszek Demkowicz (co-supervisor), Dr. Robert Moser (co-supervisor),\vspace{-0.3cm}\\ Dr. Tom Hughes, Committee Member 2, Committee Member 3}

\title{A DPG method for compressible flow problems}

\begin{document}
\maketitle

Over the last three decades, Computational Fluid Dynamics (CFD) simulations have become commonplace as a tool in the engineering and design of high-speed aircraft.  Experiments are often complemented by computational simulations, and CFD technologies have proved very useful in both the reduction of aircraft development cycles, and in the simulation of conditions difficult to reproduce experimentally.  Great advances have been made in the field since its introduction, especially in areas of meshing, computer architecture, and solution strategies.  Despite this, there still exist many computational limitations in existing CFD methods; in particular, reliable higher order and $hp$-adaptive methods for the Navier-Stokes equations that govern compressible flow.

Solutions to the Navier-Stokes equations can display shocks and boundary layers, which are characterized by localized regions of rapid change and high gradients.  The use of adaptive meshes is crucial in such settings --- good resolution for such problems under uniform meshes is computationally prohibitive and impractical for most physical regimes of interest.  However, the construction of ``good" meshes is a difficult task, usually requiring a-priori knowledge of the form of the solution.  An alternative to such is the construction of automatically adaptive schemes; such methods begin with a coarse mesh and refine based on the minimization of error.  However, this task is difficult, as the convergence of numerical methods for problems in CFD is notoriously sensitive to mesh quality.  Additionally, the use of adaptivity becomes even more difficult in the context of higher order and $hp$ methods \cite{BoeingHigherOrder}.

Many of the above issues are tied to the notion of \emph{robustness}, which we define loosely for CFD applications as the degradation of the quality of numerical solutions on a coarse mesh with respect to the Reynolds number, or nondimensional viscosity. For typical physical conditions of interest for the compressible Navier-Stokes equations, the Reynolds number dictates the scale of shock and boundary layer phenomena, and can be extremely high --- on the order of 1e7 in a unit domain.  For an under-resolved mesh, the Galerkin finite element method develops large oscillations which prevent convergence and pollute the solution.

The issue of robustness for finite element methods was addressed early on by Brooks and Hughes in the SUPG method \cite{SUPG}, which introduced the idea of residual-based stabilization to combat such oscillations. Such methods have been widely used in industry; however, these methods are limited to low-order approximations. Additionally, such methods are typically formulated for model problems and then applied to the Navier-Stokes equations, where the extrapolation can be inexact and ad-hoc.

Residual-based stabilizations can alternatively be viewed as modifying the standard finite element test space, and consequently the norm in which the finite element method converges. Demkowicz and Gopalakrishnan generalized this idea in 2009 by introducing the discontinous Petrov-Galerkin (DPG) method with optimal test functions, where test functions are determined such that they minimize the discrete linear residual in a dual space.  Under the ultra-weak variational formulation, these test functions can be computed locally to yield a symmetric, positive-definite system.  The DPG method was quickly used to solve the convection and convection-diffusion problems \cite{DPG1,DPG2,DPG3}, and has recently been shown to yield a robust higher order adaptive method for the convection-diffusion model problem in arbitrary dimensions \cite{DPGrobustness,DPGrobustness2}.

\paragraph{Goal:} Our goal is to develop a DPG method and higher order adaptive scheme for the steady state compressible laminar Navier-Stokes equations in transonic/supersonic regimes that is robust over a range of Mach and Reynolds numbers. In particular, we hope to present a method for which automatic $hp$-adaptivity can be applied to problems in compressible flow.

\subsection*{Area A: Applicable Mathematics}

The main theoretical thrust of this research is to develop a DPG method that is provably robust for singular perturbation problems in CFD, but does not suffer from discretization error in the approximation of test functions \cite{DPGrobustness, DPGrobustness2}.  I will develop such a method for the prototypical singular perturbation problem of convection-diffusion, demonstrate that the method does not suffer from error in the approximation of test functions, and prove the robust bound of the $L^2$ error by the energy error in which DPG is optimal.  In doing so, I will employ a new inflow boundary condition which, compared to standard Dirichlet inflow conditions, induces more well-behaved solutions to the adjoint problem, and describe the effect upon the robustness of the DPG method.  I will then extend this method to the linearized Navier-Stokes equations, and conduct a similar analysis of the robustness of DPG for the linearized Navier-Stokes system.

\subsection*{Area B: Numerical Analysis and Scientific Computation}

A large thrust of the numerical work here has focused on the development of a 2D compressible flow code under the parallel DPG Camellia codebase, developed by Nathan Roberts at ICES \cite{Camellia}.  In particular, we have developed a framework allowing for rapid implementation of problems and the easy application of higher order and $hp$-adaptive schemes based on a natural error representation function that stems from the DPG residual \cite{DPG2, DPG3}. Additionally, the effectiveness and necessity of anisotropic refinement schemes for problems in CFD has been demonstrated several times in the literature \cite{anisotropy1,anisotropy2}. The DPG error representation function has been shown to yield an effective and natural vector-valued residual with which to drive refinement, which we hope to generalize to anisotropic adaptive schemes.

A recent theoretical development has been a Hessian-based DPG method that aims to minimize the \emph{nonlinear} residual in the dual space.  This method will augment the positive definite DPG stiffness matrix with a symmetric update determined by the form of the Hessian and the nonlinear residual.  I will implement such a nonlinear approach and study its effect on the convergence of the nonlinear steady-state problem and the feasibility of different solvers for the new discrete system.

Finally, the DPG method under the ultra-weak formulation yields itself very naturally to a host of parallelizable solvers.  In particular, I will implement the distributed Schur complement method and study its scalability.

\subsection*{Area C: Mathematical Modeling and Applications}

I will apply the DPG method to a several convection diffusion problems which mimic difficult problems in compressible flow simulations, including boundary layer and closed streamline problems.  I will also solve a viscous Burgers' equation as an extension of DPG to nonlinear problems.  Finally, the effectiveness of DPG as a numerical method for compressible flow will be assessed with the application of DPG to common transonic and superonic benchmark problems.  In particular, I will use DPG to solve the flat plate problem, the ramp problem, flow over a Gaussian bump, and flow over an airfoil over a range of Mach numbers and laminar Reynolds numbers.

Time permitting, I will also explore the regularization of the inviscid convection and Euler equations using the full convection-diffusion and Navier-Stokes equations in the small-viscosity limit.

%Additionally, we will seek to pursue the application of DPG to the inviscid Euler equations

### 2012-06-20 docs/Jesse/Proposal/problemRange.tex

The linearized Navier-Stokes equations are the result of small perturbation assumptions applied to the full Navier-Stokes equations. Under such assumptions, the flow is largely uniform, with slight variations that are small compared to the uniform stream velocity. We are interested in the linearized Navier-Stokes equations mainly for mathematical purposes;

@@

The Burgers' equation is physically derived from the incompressible Navier-Stokes equations under the assumption that $\grad p \approx 0$, or that the pressure field is near constant. A feature of the Burgers' equation not present in convection-diffusion is that, due to the presence of the nonlinear term, it can develop shock discontinuities in its solutions in finite time.

The Burgers' equation has also been used to study the phenomenon of turbulence; however, the Burgers equation does not exhibit the chaotic nature and sensitivity to initial conditions that characterizes turbulence as observed in the full and incompressible Navier-Stokes equations. In the scope of this dissertation, Burgers shall be used as to test the extension of our numerical method to nonlinear problems.

@@

The Euler equations are a result of neglecting the viscous terms in the Navier-Stokes equations. Physically speaking, these assumptions correspond to the inviscid limit, as well as a particular class of boundary conditions (for example, the wall boundary condition $\boldsymbol u = 0$ does not hold in the inviscid limit).

\todo{express coupling - Drela's IBL work} \cite{BoeingDrela}
\todo{Express that mathematically, the Euler equations are difficult}

>>>>

The linearized Navier-Stokes equations are the result of small perturbation assumptions applied to the full Navier-Stokes equations. Under such assumptions, the flow is largely uniform, with slight variations that are small compared to the uniform stream velocity. We are interested in the linearized Navier-Stokes equations mainly for mathematical purposes - as the solution to the full Navier-Stokes equations involves a series of solutions for linearized Navier-Stokes, we wish to investigate the behavior of our numerical method with respect to this system.

@@

The Burgers' equation is physically derived from the incompressible Navier-Stokes equations under the assumption that $\grad p \approx 0$, or that the pressure field is near constant. A feature of the Burgers' equation not present in convection-diffusion is that, due to the presence of the nonlinear term, it can develop shock discontinuities in its solutions in finite time.  The Burgers' equation has also been used to study the phenomenon of turbulence; however, the Burgers equation does not exhibit the chaotic nature and sensitivity to initial conditions that characterizes turbulence as observed in the full and incompressible Navier-Stokes equations. In the scope of this dissertation proposal, Burgers shall be used as to test the extension of our numerical method to nonlinear problems.

@@

The pure convection equation is a result of neglecting the viscous term in the convection-diffusion equation. Physically speaking, these assumptions correspond to the inviscid limit, as well as a particular class of boundary conditions (for example, the wall boundary condition $u = 0$ and prescribed inflow condition may be incompatible in the inviscid limit).  The Euler equations are likewise a result of neglecting the viscous terms in the Navier-Stokes equations.  However, these problems can be ill-posed in the continuous setting.  Take, for example, the vortex problem in Figure~\ref{fig:convCirc}.  A feature of the convection equation is that there is no crosswind diffusion - thus, materials do not mix across streamlines.  However, for the vortex problem, this also implies that the solution on any closed streamline can take any arbitrary value, and is thus undefined.

\begin{figure}[!h]
\centering
\includegraphics[scale = .22]{figs/convCirc.png}
\caption{Setup for the vortex problem.}
\label{fig:convCirc}
\end{figure}
Formally speaking, the solution to the vortex problem is taken to be the solution to the convection-diffusion equation (with appropriate outflow boundary conditions) as the viscosity tends towards zero, in which case, the solution in the interior would be uniformly zero.  This motivates the need for \emph{artificial viscosity} methods with which to regularize inviscid solutions.  The topic is expansive, and we direct the reader towards \cite{Barter} for a more detailed discussion of past and present artificial viscosity methods.

There is still much interest in solving the inviscid Euler equations in industry --- the full Navier-Stokes models often prove difficult to solve due to the mathematical nature of the equations, and the coupling of the Euler equations with boundary layer models has been successful in simulating many phenomena in compressible flow \cite{BoeingDrela}.
